Simulations of gravity-driven flow of binary liquids in microchannels

### 2011-04-06 Review/answers.tex
Finished all the revisions and submitted it to the journal.

\textbf{Answer:} The reviewer is right that accuracy of the method deteriorate with the high
viscosity ratio. However, the parameters we chose here as $\tau_{\mathrm{liq}}=3.5-7.5$ and with
grids in the vertical direction as $N_y=202$ is sufficient to suppress effects. We have added all
the necessary references in the text, lines $481-491$.

@@

\textbf{Answer:} We redrew all the figures. For the figure 8 the data was extracted from the
figure and rescaled for the comparison reasons.

@@

\textbf{Question:}
\begin{quotation}
The use of the variable $Ca_{lit}$ and the calculation of the corresponding pressure gradient yield
estimations of the body force to apply and of the film thickness to choose grid resolution. The
presentation
of the corresponding procedure is repeated twice in Section 5-1 for $Ca=0.005$ and $Ca=0.05$. It is
again described in a diﬀerent way in Section 5-4. Similar equations for the calculation of the
pressure
gradient are thus given in Equations 6, 14 and 16. The Authors should summarize these equations
into one unless they explain their differences. In fact, it would probably be straightforward not to
detail the calculation of estimated variables since they are intermediary variables which allow the
design of the benchmark but do not inﬂuence directly the results. Also, the Authors could possibly
say they control and impose directly the body force without describing all the estimation procedure.
\end{quotation}

\textbf{Answer:} Thank you for the suggestion. We agree that there is redundant information
about initialization and this was taken into account by reducing repetitive text and
introducing step-by-step explations, see Section about initialization (lines 259-305). We also added
the
recommendations for controlling parameters in the same section.  The
dimensionalization of the problem and proper benchmarking is quite difficult in the lattice
Boltzmann framework. You can refer to the lbmethod.org/forum where half of all questions related to
the lattice Boltzmann method are refered to the non-dimensionalization procedure. Therefore, we
decided to
keep suggestions for controlling and initialization of simulations as informative guidelines for the
LBM community.

@@

\textbf{Answer:} The factor $1/8$ is taken for the maximum velocity for the Poiseille flow between
plates. Our assumption that the bubble travels with the constant velocity which equals to the
maximum velocity of the Poiseuille profile between plates. Factor $1/12$ comes from the averaging
of the flow rate. However, bubble moves faster than the liquid, i.e. the average liquid velocity.
That's why the factor $1/8$ was chosen. The simulations show that factor $1/8$ works reasonably
well.

@@

\textbf{Answer:} We wrote an additional paragraph emphasizing the connection of the physical world
parameters with the lattice Botlzmann parameters (lines $149-160$). As well more explanations are
put towards the
interface surface tension description (lines $136-139$). The reference to the
continuity and the Navier-Stokes equations are added (lines $140-143$). The equation (2) is not
rewritten in terms of
generic variables as far as it's a historical LBM notation for the Navier-Stokes and phase
governing equations. The notations have been corrected.

@@

\textbf{Answer:} We added a small steady-state study (lines $334-341$). We added an additional
paragraph which shows the change of flow pattern for the
streamlines as it is indicated in literature (lines $463-471$, Figure 8).

>>>>

\textbf{Answer:} The reviewer is right that accuracy of the method deteriorates with the high
viscosity ratio. However, the parameters we chose here as $\tau_{\mathrm{liq}}=3.5-7.5$ and with
grids in the vertical direction as $N_y=202$ is sufficient to suppress effects. We have added all
the necessary references in the text, lines $481-491$.

@@

\textbf{Answer:} We redrew all the figures. For the figure 8 the data was extracted from the
figure and rescaled for comparison reasons.

@@

\textbf{Question:}
\begin{quotation}
The use of the variable $Ca_{lit}$ and the calculation of the corresponding pressure gradient yield
estimations of the body force to apply and of the film thickness to choose grid resolution. The
presentation
of the corresponding procedure is repeated twice in Section 5-1 for $Ca=0.005$ and $Ca=0.05$. It is
again described in a different way in Section 5-4. Similar equations for the calculation of the
pressure
gradient are thus given in Equations 6, 14 and 16. The Authors should summarize these equations
into one unless they explain their differences. In fact, it would probably be straightforward not to
detail the calculation of estimated variables since they are intermediary variables which allow the
design of the benchmark but do not inﬂuence directly the results. Also, the Authors could possibly
say they control and impose directly the body force without describing all the estimation procedure.
\end{quotation}

\textbf{Answer:} Thank you for the suggestion. We agree that there is redundant information
about initialization and this was taken into account by reducing repetitive text and
introducing step-by-step explanations, see Section about initialization (lines $259-305$). We also
added
the
recommendations for controlling parameters in the same section.  The
dimensionalization of the problem and proper benchmarking is quite difficult in the lattice
Boltzmann framework. You can refer to the lbmethod.org/forum where half of all questions related to
the lattice Boltzmann method are refered to the non-dimensionalization procedure. Therefore, we
decided to
keep suggestions for controlling and initialization of simulations as informative guidelines for the
LBM community.

@@

\textbf{Answer:} The factor $1/8$ in Eq. (5) is taken for the maximum velocity for the Poiseille
flow between
plates. Our assumption is that the bubble travels with the constant velocity which equals to the
maximum velocity of the Poiseuille profile between plates. Factor $1/12$ comes from the averaging
of the flow rate. However, bubble moves faster than the liquid, i.e. the average liquid velocity.
That is why the factor $1/8$ was chosen. The simulations show that factor $1/8$ works reasonably
well.

@@

\textbf{Answer:} We wrote an additional paragraph emphasizing the connection of the
physical parameters with the lattice Botlzmann parameters (lines $149-160$). As well more
explanations are
put towards the
interface surface tension description (lines $136-139$). The reference to the
continuity and the Navier-Stokes equations are added (lines $140-143$). The equation (2) is not
rewritten in terms of
generic variables as far as it's a historical LBM notation for the Navier-Stokes and phase
governing equations. The notations have been corrected.

@@

\textbf{Answer:} We added a steady-state study (lines $334-341$). We added an additional
paragraph which shows the change of flow pattern for the
streamlines as it is indicated in literature (lines $463-471$, Figure 8).

### 2011-04-06 Text/paper.tex
Finished all the revisions and submitted it to the journal.

The binary liquid free-energy LB model due to \citet{swift} we used
simulates two liquids with the assumption of uniform overall
density. While the classical Bretherton problem is stated for gas and liquid, which are of
significantly
different densities and viscosities, it was indicated \cite{bretherton} that inertia effects were
negligible. Moreover, the results of \citet{giavedoni-numerical} and \citet{heil-bretherton} show
{\color{red} insignificant} Reynolds number ($\Re$) effects on the film thickness for a relatively wide
range of Reynolds
numbers. 
{\color{red} For example, \citet{giavedoni-numerical} suggested that the Reynolds number effects
are negligible for $\Ca\leq0.05$ and have moderate impact for $\Ca>0.05$ in the range of Reynolds
number from $0$ to $70$. Later on, \citet{heil-bretherton} extended these results up to
$\Re = 300$. This indicates that while the Reynolds number influence on the established film thickness is insignificant
($7$ percent from the film thickness measured at $\Re=0$), the pressure distribution and the flow field near the front bubble tip
can be affected by changes in $\Re$.  Since in the present work we only deal with flows with $\Re < 20$, we can safely
neglect any inertia effects and conclude that the major governing parameter for microchannel Bretherton flows is not the density ratio,
but the
viscosity ratio}. Moreover, the results based on uniform density, as in the case of the LBM
binary liquid model, are in good agreement with other simulations
\cite{giavedoni-numerical,heil-bretherton} in which inertia effects were taken into account.

@@

{\color{red}
The level set method is also represented by a number of works \cite{fukugata-levelset,
lakehal-micro, carlson-comparison}. While the bubble shapes  were found to be in a good agreement
with the experiments, the range of studied bubble lengths is confined to $2-3$ channel diameters.
Thus, the bubble radius changes significantly and cannot be used as a classical Bretherton problem
benchmark. However, the works provide insight to the bubble formation and to the way small bubbles
flow in the
microchannels.}

\section{Lattice Boltzmann binary liquid model}
The lattice Boltzmann equation (LBE) operates on a rectangular grid representing the
physical domain. It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density and momentum. LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another along some 
directions specified by the discrete velocity set.
The LBE is typically implemented as follows:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)=\omega f_i^{eq}(\bm{x},t)-(1-\omega) f_i(\bm{x},t) +
F_i,&&\text{ collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $f_i$ is the probability distribution function in the direction $\bm{c_i}$,
{\color{red} $f_i^{eq}$ is the equilibrium probability distribution function}, $\omega$ is the
relaxation parameter, and $F_i$ is the external force population. {\color{red} The force population
is represents an texternal physical force and is implemented in the current work using the scheme
outlined in \citet{guo}.}

@@

The model we use is a two-dimensional nine-velocity (D2Q9) model,
with equilibrium populations \cite{pooley-contact}:
\begin{equation}
\label{set:equilibrium:binary}
\begin{aligned}
&f_i^{eq}&&=w_i 
\biggl(3
p_0 - k \phi \Delta \phi
+\rho\frac{u_{\alpha}c_{i\alpha}}{c_s^2}+\rho \frac{Q_{i\alpha\beta}u_{\alpha } u_ {
\beta}}{2 c_s^4}\biggr)\\
&&&+k\bigl(w_i^{xx} (\partial_x \phi)^2+w_i^{yy} (\partial_y \phi)^2 +w_i^{xy} \partial_x
\phi \partial_y \phi \bigr), 1\leq i \leq 8\\
&f_0^{eq}&&=\rho-\sum_{i\neq0}{f_i^{eq}}\\
&g_i^{eq}&&=w_i\left(\Gamma \mu + \phi\frac{ c_{i\alpha} u_{i\alpha}}{c_s^2}+\phi
\frac{Q_{i\alpha\beta}u_{\alpha}u_{\beta}}{2 c_s^4}\right), 1\leq i \leq 8 \\
&g_0^{eq}&&=\phi-\sum_{i\neq0}{g_i^{eq}}\quad,
\end{aligned}
\end{equation}
where $\Gamma$ is the mobility parameter; the chemical potential
$\mu=-A\phi+A\phi^3-k\Delta\phi$; $k$ is the parameter related to the surface
tension; $A$ is the parameter of the free-energy model. The bulk pressure
is expressed as $p_0=c_s^2 \rho +A (-0.5 \phi^2+0.75 \phi^4)$ {\color{red} with
the sound speed $c_s^2=1/3$}. 
Parameters specific to the D2Q9 grid are the weights
$w_i=\left\{\frac{4}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},
\frac{1}{36},\frac{1}{36},\frac{1}{36},\frac{1}{36}\right\}$, and the tensor
$Q_{i\alpha\beta}=c_{i\alpha} c_{i\beta} - c_s^2 \delta_{\alpha\beta}$.  
{\color{red}
Other weights 
%related to the
%inclusion of the surface tension coefficient into the equations 
are as follows:}
$w^{xx}_{1-2}=w^{yy}_{3-4}=1/3$, $w^{xx}_{3-4}=w^{yy}_{1-2}=-1/6$,
$w^{xx}_{5-8}=w^{yy}_{5-8}=-1/24$, $w^{xy}_{1-4}=0$, $w^{xy}_{5-6}=1/4$ and
$w^{xy}_{7-8}=-1/4$. The set of equations (\ref{set:equilibrium:binary}) restores the
macroscopic
fluid equations as:
\begin{equation}
\begin{aligned}
&\partial_t \rho+ \partial_{\alpha} \rho u_{\alpha}=0\\
&\rho\left(\partial_t+u_{\beta}\partial_{\beta}\right) u_{\alpha}= F_{\alpha}
-\partial_{\beta}P_{\alpha \beta} +
\nu\partial_{\beta}\left(\partial_{\alpha}u_{\beta}+\partial_{\beta} u_{\alpha}\right)\\
&\partial_t \phi + \partial_{\alpha} \phi u_{\alpha}=M \partial^2_{\beta\beta} \mu,
\end{aligned}
\label{binary:fluid:system}
\end{equation}
where $\nu=c_s^2 (\tau-1/2)$ is the viscosity,
$M=\Gamma(\tau_{\phi}-1/2)$ is the mobility parameter, and $\tau=\frac{1}{\omega}$ and $\tau_{\phi}$
are the relaxation parameters of density and phase fields, {\color{red}
$P_{\alpha\beta}=\Bigl(p_0-k\phi \Delta \phi -\frac{k}{2}|\nabla \phi|^2\Bigr)\delta_{\alpha\beta}
+ k \partial_{\alpha} \phi \partial_{\beta} \phi$  \cite{pooley-contact}.The interface tension value
in the framework of the binary liquid model is $\gamma=\sqrt{\frac{8 k
A}{9}}$. The inclusion of the interface tension to the momentum flux tensor is done through the
coefficients $k$, $A$ and weights $w_i^{\alpha\beta}$.

@@

{\color{red}
While the lattice Boltzmann system has parameters such as the surface tension, the gas and liquid
viscosities, etc., those parameters are not the representative and
proportional quantities of the parameters in a physical system. The parameters of the lattice
Boltzmann are connected with the physical world's parameters only through the non-dimensional
numbers governing the physics of the problem. In our case, these numbers are the capillary number $\Ca$ and the
viscosity ratio $\frac{\mu_{\mathrm{liq}}}{\mu_{\mathrm{gas}}}$, which are
obtained from the physical world and then matched through the lattice Boltzmann quantities.
 The set of
the fluid and phase equations \eqref{binary:fluid:system} is valid in the lattice Boltzmann space
and in the physical domain. Therefore, one can substitute any quantity, i.e.
$U_{\mathrm{bubble}}$, in the physical units or in the lattice Boltzmann units as soon as the
capillary number is the same in both worlds.}

@@

{\color{red}
For demonstration purposes we present the simulation parameters choice for $\Ca=0.05$.
In these
examples we use $k=0.04$ and $A=0.04$ as optimal binary liquid model parameters.
That implies the surface tension to be $\sqrt{\frac{8 k
A}{9}}=0.0377$ and the characteristic width of the interface
$5\sqrt{\frac{k}{A}}=5$ lattice Boltzmann units (grid nodes). For the sake of simplicity and
stability, the relaxation times for liquid and gas phases are taken as
$\tau_{\mathrm{liq}}=2.5$ and $\tau_{\mathrm{gas}}=0.7$, which correspond
to the liquid over gas kinematic viscosity ratio equal to $10$.

@@

where $\phi_0$ is the grid coordinate where phase field is $0$, $N_y-2$

@@

multiphase models to the best authors' knowledge is not yet derived. Thus, the simulation grid size

@@

flow is close to the Poiseuille flow:

@@

$100000-300000$ time steps due to reach steady-state, see Section \ref{sec:steady:state}.

@@

{\color{red}
Note that the initialization procedure is a good guide of designing the simulations. For example
for the grid refinement, Section \ref{section:grid:refinement}, we keep the same velocity
$U_{\mathrm{bubble}}$ and $\Ca$. This implies the conservation of the quantity
$H_{\mathrm{eff}}^2\frac{\mathrm{d}P}{\mathrm{d}x}=(N_y-2)^2\frac{\mathrm{d}P}{\mathrm{d}
x } = \mathrm{const}$. For designing capillary range simulations, Section
\ref{section:capillary:range}, we use a proportionality law to initialize body forces knowing
results of just one simulation:
\begin{equation}
\begin{aligned}
&\Ca_{\mathrm{lit}} \propto U_{\mathrm{bubble}}\\
&U_{\mathrm{bubble}} \propto \frac{\mathrm{d}P}{\mathrm{d}x} H_{\mathrm{eff}}^2\\
&\Ca_{\mathrm{lit}} \propto \frac{\mathrm{d}P}{\mathrm{d} x} H_{\mathrm{eff}}^2 \text{ or }\\
&\frac{\mathrm{d}P}{\mathrm{d} x} \propto \frac{\Ca_{\mathrm{lit}}}{H_{\mathrm{eff}}^2},
\end{aligned}
\end{equation}
where the subscript ,,lit'' stands for the predicted capillary number
\cite{giavedoni-numerical,heil-bretherton}. 
}
{\color{red}
\subsection{Steady state}
\label{sec:steady:state}
The simulations were performed in order to determine the number of iterations for the bubble to
reach the steady state. The grid was chosen as $202 \times 3000$. The body force was $0.375\times
10^{-6}$. The initial film thickness was taken as $12$ nodes. Table \ref{table:steady:state}
summarizes the simulation results in terms of the film thickness in the middle of the bubble and the
velocity of the interface. We can conclude it is enough to have $100000-300000$ iterations for
the bubble to approach the steady state.
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|X|X|X|}
\hline
$N_\mathrm{iter}$&$\delta$&$U_{\mathrm{bubble}}$&{\color{red} $\Re$}\\
\hline
$10000$&$ 0.0655$&$0.0027$&$0.81$\\
\hline
$14000$&$0.0608$&$0.0027$&$0.81$\\
\hline
$18000$&$0.0577$&$0.0027$&$0.81$\\
\hline
$22000$&$0.0630$&$0.0027$&$0.81$\\
\hline
$26000$&$0.0660$&$0.0027$&$0.81$\\
\hline
$30000$&$0.0661$&$0.0027$&$0.81$\\
\hline
\end{tabularx}
\caption{The results for steady-state calculations. The simulated domain is
of size $202 \times 3000$. $U_{\mathrm{bubble}}$ is the velocity of the interface at the fron tip of
the bubble. $\Ca=0.05$ was taken to initialize the simulation. One can see that in terms of
velocities and film thicknesses $100000-300000$ iterations are enough.
\label{table:steady:state}}
\end{table}
}

@@

\subsection{The film variation over the bubble}
\label{section:film:thickness:variation}
Next, we investigated the variation of the film thickness over the bubble length.
The comparison for the thickness variation is presented in Fig.~\ref{fig:sehgal:bubble:length}.
\begin{figure}
\includegraphics[width=0.5\textwidth]{Figures/Bubble/bubble_sehgal_new.eps}\hfill
\includegraphics[width=0.5\textwidth]{Figures/Bubble/bubble_length_new.eps}\\
\caption{A qualitative comparison for the film thickness across
the bubble length. {\color{red} Left} (courtesy of \citet{sehgal-microchannel}),
{\color{red} Right}  (present simulations); $x$ is scaled to $H_{\mathrm{eff}}$
and increases in the flow direction.{\color{red} The data from
\cite{sehgal-microchannel} was digitally acquired by the program ``Engauge Digitizer'' and scaled
to compare with the present simulations.} \label{fig:sehgal:bubble:length}}
\end{figure}
One can see a qualitative agreement, i.e. the thickness increases towards the front meniscus and
rapidly decreases towards the rear meniscus. This shape is sometimes referred as a "bullet"
shape.
{\color{red} The shape of the bubble is an important quantity, as the bubble tip shape mainly
influences the film thickness at infinity \cite{bretherton}. It was indicated by
\citet{heil-bretherton} that the inertia effects influence insignificantly the film thickness at
infinity but influence significantly the velocity pattern in front of the bubble. Therefore, to
properly describe the Bretherton phenomena the
bubble shape along with velocity pattern need to be examined.}

@@

it is enough for bubble to have a length of two-three tube diameters to apply analysis for very long
bubbles. In comparison with the axisymmetric circular geometry, for three-dimensional square shaped

@@

{\color{red} For the completeness of the presented data we also indicate the velocity streamlines
profiles, Fig. \ref{fig:streamlines:pattern}. The pattern shows that there is a transition of
having vortex in the front of the bubble $\Ca=0.22$ and not having it for larger capillary numbers
$\Ca=1.00$. \citet{giavedoni-numerical} studied the behavior of the detaching of the vortex for the
Bretherton problem. They indicate critical capillary numbers as $\Ca=0.73$ and $\Ca=0.89$ for
detaching and full vortex disappearance. \citet{heil-bretherton} indicates that the behavior of the
pattern strongly depends on the Reynolds number. The scope of this work is to indicate the
qualitative capture of the effects.  
}
\begin{figure}
\includegraphics[angle=90,width=\textwidth]{Figures/Capillary/ca022_velocity_new.eps}\\
\includegraphics[angle=90,width=\textwidth]{Figures/Capillary/ca100_velocity_new.eps}
\caption{The streamlines plots for $\Ca=0.22$ (top) and $\Ca=1.00$ (bottom). One can see the change
of pattern as indicated in works of \citet{heil-bretherton} and \citet{giavedoni-numerical}. For
smaller capillary numbers there exists a vortex in front of the bubble. For the larger once there
is none. The streamlines are presented in the reference frame moving with the bubble
interface.\label{fig:streamlines:pattern}}
\end{figure}

>>>>

The binary liquid free-energy LB model due to \citet{swift} we used
simulates two liquids with the assumption of uniform overall
density. While the classical Bretherton problem is stated for gas and liquid, which are of
significantly
different densities and viscosities, it was indicated \cite{bretherton} that inertia effects were
negligible. Moreover, the results of \citet{giavedoni-numerical} and \citet{heil-bretherton} show
{\color{red} insignificant} Reynolds number ($\Re$) effects on the film thickness for a relatively wide
range of Reynolds
numbers. 
{\color{red} For example, \citet{giavedoni-numerical} suggested that the Reynolds number effects
are negligible for $\Ca\leq0.05$ and have moderate impact for $\Ca>0.05$ in the range of Reynolds
number from $0$ to $70$. Later on, \citet{heil-bretherton} extended these results up to
$\Re = 300$. They indicate that while the Reynolds number influence on the established film
thickness is insignificant
($7$ percent from the film thickness measured at $\Re=0$), the pressure distribution and the flow field near the front bubble tip
can be affected by changes in $\Re$.  Since in the present work we only deal with flows with $\Re < 20$, we can safely
neglect any inertia effects and conclude that the major governing parameter for microchannel Bretherton flows is not the density ratio,
but the
viscosity ratio}. Moreover, the results based on uniform density, as in the case of the LBM
binary liquid model, are in good agreement with other simulations
\cite{giavedoni-numerical,heil-bretherton} in which inertia effects were taken into account.

@@

{\color{red}
The level set method is also represented by a number of works \cite{fukugata-levelset,
lakehal-micro, carlson-comparison}. While the bubble shapes  were found to be in a good agreement
with the experiments, the range of studied bubble lengths is limited to $2-3$ channel diameters.
Thus, the bubble radius changes significantly and cannot be used as a classical Bretherton problem
benchmark. However, the works provide insight to the bubble formation and to the way small bubbles
flow in microchannels.}

\section{Lattice Boltzmann binary liquid model}
The lattice Boltzmann equation (LBE) operates on a rectangular grid representing the
physical domain. It utilizes
probability distribution functions (also known as particle populations)
containing information about
macroscopic variables, such as fluid density and momentum. LBE consists of
two parts: a local collision step, and a propagation step which transports
information from one node to another along some 
directions specified by the discrete velocity set.
The LBE is typically implemented as follows:
\begin{equation}
\label{standard:implementation}
\begin{aligned}
&f_i^{*}(\bm{x},t)=\omega f_i^{eq}(\bm{x},t)-(1-\omega) f_i(\bm{x},t) +
F_i,&&\text{ collision step}\\
&f_i(\bm{x}+\bm{c_i},t+1)=f_i^{*}(\bm{x},t),&&\text{ propagation step}, 
\end{aligned}
\end{equation}
where $f_i$ is the probability distribution function in the direction $\bm{c_i}$,
{\color{red} $f_i^{eq}$ is the equilibrium probability distribution function}, $\omega$ is the
relaxation parameter, and $F_i$ is the external force population. {\color{red} The force population
represents an external physical force and is implemented in the current work using the scheme
outlined in \citet{guo}.}

@@

The model we use is a two-dimensional nine-velocity (D2Q9) model,
with equilibrium populations \cite{pooley-contact}:
\begin{equation}
\label{set:equilibrium:binary}
\begin{aligned}
&f_i^{eq}&&=w_i 
\biggl(3
p_0 - k \phi \Delta \phi
+\rho\frac{u_{\alpha}c_{i\alpha}}{c_s^2}+\rho \frac{Q_{i\alpha\beta}u_{\alpha } u_ {
\beta}}{2 c_s^4}\biggr)\\
&&&+k\bigl(w_i^{xx} (\partial_x \phi)^2+w_i^{yy} (\partial_y \phi)^2 +w_i^{xy} \partial_x
\phi \partial_y \phi \bigr), 1\leq i \leq 8\\
&f_0^{eq}&&=\rho-\sum_{i\neq0}{f_i^{eq}}\\
&g_i^{eq}&&=w_i\left(\Gamma \mu + \phi\frac{ c_{i\alpha} u_{i\alpha}}{c_s^2}+\phi
\frac{Q_{i\alpha\beta}u_{\alpha}u_{\beta}}{2 c_s^4}\right), 1\leq i \leq 8 \\
&g_0^{eq}&&=\phi-\sum_{i\neq0}{g_i^{eq}}\quad,
\end{aligned}
\end{equation}
where $\Gamma$ is the mobility parameter; the chemical potential
$\mu=-A\phi+A\phi^3-k\Delta\phi$; $k$ is the parameter related to the surface
tension; $A$ is the parameter of the free-energy model. The bulk pressure
is expressed as $p_0=c_s^2 \rho +A (-0.5 \phi^2+0.75 \phi^4)$ {\color{red} with
the sound speed $c_s^2=1/3$}. 
Parameters specific to the D2Q9 grid are the weights
$w_i=\left\{\frac{4}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},\frac{1}{9},
\frac{1}{36},\frac{1}{36},\frac{1}{36},\frac{1}{36}\right\}$, and the tensor
$Q_{i\alpha\beta}=c_{i\alpha} c_{i\beta} - c_s^2 \delta_{\alpha\beta}$.  
{\color{red}
Other weights 
%related to the
%inclusion of the surface tension coefficient into the equations 
are as follows:}
$w^{xx}_{1-2}=w^{yy}_{3-4}=1/3$, $w^{xx}_{3-4}=w^{yy}_{1-2}=-1/6$,
$w^{xx}_{5-8}=w^{yy}_{5-8}=-1/24$, $w^{xy}_{1-4}=0$, $w^{xy}_{5-6}=1/4$ and
$w^{xy}_{7-8}=-1/4$. The set of equations (\ref{set:equilibrium:binary}) restores the
macroscopic
fluid equations as:
\begin{equation}
\begin{aligned}
&\partial_t \rho+ \partial_{\alpha} \rho u_{\alpha}=0\\
&\rho\left(\partial_t+u_{\beta}\partial_{\beta}\right) u_{\alpha}= F_{\alpha}
-\partial_{\beta}P_{\alpha \beta} +
\nu\partial_{\beta}\left(\partial_{\alpha}u_{\beta}+\partial_{\beta} u_{\alpha}\right)\\
&\partial_t \phi + \partial_{\alpha} \phi u_{\alpha}=M \partial^2_{\beta\beta} \mu,
\end{aligned}
\label{binary:fluid:system}
\end{equation}
where $\nu=c_s^2 (\tau-1/2)$ is the viscosity,
$M=\Gamma(\tau_{\phi}-1/2)$ is the mobility parameter, and $\tau=\frac{1}{\omega}$ and $\tau_{\phi}$
are the relaxation parameters of density and phase fields, {\color{red}
$P_{\alpha\beta}=\Bigl(p_0-k\phi \Delta \phi -\frac{k}{2}|\nabla \phi|^2\Bigr)\delta_{\alpha\beta}
+ k \partial_{\alpha} \phi \partial_{\beta} \phi$  \cite{pooley-contact}.The interface tension value
in the framework of the binary liquid model is $\gamma=\sqrt{\frac{8 k
A}{9}}$. The inclusion of the interface tension in the momentum flux tensor is done through the
coefficients $k$, $A$ and weights $w_i^{\alpha\beta}$.

@@

{\color{red}
While the lattice Boltzmann system has parameters such as the surface tension, the gas and liquid
viscosities, etc., those parameters are not the representative and
proportional quantities of the parameters in a physical system. The parameters of the lattice
Boltzmann are connected with the physical parameters only through the non-dimensional
numbers governing the physics of the problem. In our case, these numbers are the capillary number $\Ca$ and the
viscosity ratio $\frac{\mu_{\mathrm{liq}}}{\mu_{\mathrm{gas}}}$, which are
obtained from the physical world and then matched through the lattice Boltzmann quantities.
 The set of
the fluid and phase equations \eqref{binary:fluid:system} is valid in the lattice Boltzmann space
and in the physical domain. Therefore, one can substitute any quantity, i.e.
$U_{\mathrm{bubble}}$, in the physical units or in the lattice Boltzmann units as soon as the
capillary number is the same in both worlds.}

@@

{\color{red}
For demonstration purposes we present the simulation parameters choice for $\Ca=0.05$.
In these
examples we use $k=0.04$ and $A=0.04$ as numerically optimal binary liquid model parameters.
That implies the surface tension to be $\sqrt{\frac{8 k
A}{9}}=0.0377$ and the characteristic width of the interface
$5\sqrt{\frac{k}{A}}=5$ lattice Boltzmann units (grid spacings). For the sake of simplicity and
stability, the relaxation times for liquid and gas phases are taken as
$\tau_{\mathrm{liq}}=2.5$ and $\tau_{\mathrm{gas}}=0.7$, which correspond
to the liquid over gas kinematic viscosity ratio equal to $10$.

@@

where $\phi_0$ is the grid coordinate where the phase field is $0$, $N_y-2$

@@

multiphase models to the best of the authors' knowledge is not yet derived. Thus, the simulation
grid size

@@

flow is close to planar Poiseuille flow with $U_{\mathrm{bubble}}$ being the maximun in the
Poiseuille profile:

@@

$100000-300000$ time steps to reach steady-state, see Section \ref{sec:steady:state}.

@@

{\color{red}
Note that the initialization procedure is a good guide of designing the simulations. For example
for the grid refinement, Section \ref{section:grid:refinement}, we keep the same velocity
$U_{\mathrm{bubble}}$ and $\Ca$. This implies the conservation of the quantity
$H_{\mathrm{eff}}^2\frac{\mathrm{d}P}{\mathrm{d}x}=(N_y-2)^2\frac{\mathrm{d}P}{\mathrm{d}
x } = \mathrm{const}$. For designing capillary range simulations, Section
\ref{section:capillary:range}, we use a proportionality law to initialize body forces knowing
results of just one simulation:
\begin{equation}
\begin{aligned}
&\Ca_{\mathrm{lit}} \propto U_{\mathrm{bubble}}\\
&U_{\mathrm{bubble}} \propto \frac{\mathrm{d}P}{\mathrm{d}x} H_{\mathrm{eff}}^2\\
&\Ca_{\mathrm{lit}} \propto \frac{\mathrm{d}P}{\mathrm{d} x} H_{\mathrm{eff}}^2 \text{ or }\\
&\frac{\mathrm{d}P}{\mathrm{d} x} \propto \frac{\Ca_{\mathrm{lit}}}{H_{\mathrm{eff}}^2},
\end{aligned}
\end{equation}
where the subscript ,,lit'' stands for the predicted capillary number
\cite{giavedoni-numerical,heil-bretherton}. 
}
{\color{red}
\subsection{Steady state}
\label{sec:steady:state}
The simulations were performed in order to determine the number of iterations for the bubble to
reach the steady state. The grid was chosen as $202 \times 3000$. The body force was $0.375\times
10^{-6}$. The initial film thickness was taken as $12$ nodes. Table \ref{table:steady:state}
summarizes the simulation results in terms of the film thickness in the middle of the bubble and the
velocity of the interface as a function of time. We can conclude that it is enough to have
$100000-300000$ iterations for
the bubble to approach the steady state.
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|X|X|X|}
\hline
$N_\mathrm{iter}$&$\delta$&$U_{\mathrm{bubble}}$&{\color{red} $\Re$}\\
\hline
$100000$&$ 0.0655$&$0.0027$&$0.81$\\
\hline
$140000$&$0.0608$&$0.0027$&$0.81$\\
\hline
$180000$&$0.0577$&$0.0027$&$0.81$\\
\hline
$220000$&$0.0630$&$0.0027$&$0.81$\\
\hline
$260000$&$0.0660$&$0.0027$&$0.81$\\
\hline
$300000$&$0.0661$&$0.0027$&$0.81$\\
\hline
\end{tabularx}
\caption{The results for steady-state calculations. The simulated domain is
of size $202 \times 3000$. $U_{\mathrm{bubble}}$ is the velocity of the interface at the fron tip of
the bubble. $\Ca=0.05$ was taken to initialize the simulation. One can see that in terms of
velocities and film thicknesses $100000-300000$ iterations are enough.
\label{table:steady:state}}
\end{table}
}

@@

\subsection{The film variation over the bubble}
\label{section:film:thickness:variation}
Next, we investigated the variation of the film thickness over the bubble length.
The comparison for the thickness variation is presented in Fig.~\ref{fig:sehgal:bubble:length}.
\begin{figure}
\includegraphics[width=0.5\textwidth]{Figures/Bubble/bubble_sehgal_new.eps}\hfill
\includegraphics[width=0.5\textwidth]{Figures/Bubble/bubble_length_new.eps}\\
\caption{A qualitative comparison for the film thickness across
the bubble length. {\color{red} Left} (courtesy of \citet{sehgal-microchannel}),
{\color{red} Right}  (present simulations); $x$ is scaled to $H_{\mathrm{eff}}$
and increases in the flow direction.{\color{red} The data from
\cite{sehgal-microchannel} was digitally acquired by the program ``Engauge Digitizer'' and scaled
to compare with the present simulations.} \label{fig:sehgal:bubble:length}}
\end{figure}
One can see a qualitative agreement, i.e. the thickness increases towards the front meniscus and
rapidly decreases towards the rear meniscus. This shape is sometimes referred as a "bullet"
shape.
{\color{red} The shape of the bubble is an important quantity, as the bubble tip shape mainly
influences the film thickness at infinity \cite{bretherton}. It was indicated by
\citet{heil-bretherton} that inertia effects influence insignificantly the film thickness at
infinity but influence significantly the velocity pattern in front of the bubble. Therefore, to
properly describe the Bretherton phenomena the
bubble shape along with velocity pattern needs to be examined.}

@@

it is enough for bubbles to have a length of two-three tube diameters to apply analysis for very
long
bubbles. In comparison with the axisymmetric geometry, for three-dimensional square shaped

@@

{\color{red} For the completeness of the presented data we also indicate the velocity streamlines
profiles, Fig. \ref{fig:streamlines:pattern}. The pattern shows that there is a transition of
having vortex in the front of the bubble for $\Ca=0.22$ and not having it for larger capillary
numbers
$\Ca=1.00$. \citet{giavedoni-numerical} studied the behavior of the detaching of the vortex for the
Bretherton problem. They indicate critical capillary numbers as $\Ca=0.73$ and $\Ca=0.89$ for
detaching and full vortex disappearance. \citet{heil-bretherton} indicates that the behavior of the
pattern strongly depends on the Reynolds number. The scope of this work is to indicate the
qualitative capture of the effects.  
}
\begin{figure}
\includegraphics[angle=90,width=\textwidth]{Figures/Capillary/ca022_velocity_new.eps}\\
\includegraphics[angle=90,width=\textwidth]{Figures/Capillary/ca100_velocity_new.eps}
\caption{The streamlines plots for $\Ca=0.22$ (top) and $\Ca=1.00$ (bottom). One can see the change
of pattern as indicated in works of \citet{heil-bretherton} and \citet{giavedoni-numerical}. For
smaller capillary numbers there exists a vortex in front of the bubble. For the larger once there
is none. The streamlines are presented in the reference frame moving with the bubble
interface.\label{fig:streamlines:pattern}}
\end{figure}
